Experience Report: Haskell in Computational Biology

### 2012-06-27 mrfy_experience_report.tex
import hmmnode definition direct from the code
reflect new AA type
Be more honest about Viterbi: it's still slower than C++ by a lot (but the C++ version may not even be correct!)

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
When performance is paramount,
software is usually written in C~or~C++. 
When convenience, readability, and
productivity are more important,
software is usually written in a dynamically typed
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on experience using a third kind of language,
Haskell: 
\begin{itemize}
\item
We had to reimplement an
algorithm already implemented in~C++, 
but the implementation required little effort,
and the connection between code and
algorithm is unusually clear (\secref{viterbi}).
Our new Haskell code performs competitively with~C++,
and it was easy to parallelize
(\secref{perf}). 
\remark{Can we summarize the performance results?  Speedup?}
\item
Higher-order functions made it unusually easy to
create and experiment with new stochastic-search algorithms
(\secref{hofs}).
\item
Haskell slowed us down in only one way: it's hard to understand and
improve the
performance of a Haskell program (\secref{awkward-profiling}).
%%  \item
%%  The only significant impediment that Haskell presented to our
%%  research was the difficulty of understanding and improving the
%%  performance of Haskell programs (\secref{awkward-profiling}).
\item
Although authors Daniels and Gallant are computational
biologists with little background in functional programming,
we~used Haskell successfully.
And Haskell has made it easy to explore new research ideas;
by~contrast,
our group's C++ code has not been a good vehicle for exploring new ideas
(\secref{comparo}).
\item
The Haskell community offers libraries and tools that
promise powerful abstractions.
Some worked as advertised, saved us lots of effort, and were a pleasure
to~use.
Others, not so much.
And~we could not tell ahead of time which experience we were likely to have
 (\secref{penumbra}).
\end{itemize}

@@

In~idiomatic Haskell, 
we might represent an individual amino acid~$x$
using a value of algebraic data type:
\begin{smallverbatim}
data Amino = Ala | Cys | Asp | Glu | Gly | ...
\end{smallverbatim}
But our primary goal was to get something working quickly,
and our hidden Markov models are stored using a
legacy file format that represents each amino acid as a small integer.
Our representation is therefore
\begin{smallverbatim}
type AA = Int -- amino acid using HMMER coding
\end{smallverbatim}
Either way, a query sequence is (and should be) an immutable vector of
amino acids.
%type QuerySeq = Vector AA
%\end{smallverbatim}

@@

The representation of a node includes the probabilities of transitions
into the states of that node, plus the tables
of emission probabilities for the match and insertion states.
\begin{smallverbatim}
type EProbs  = Vector LogProb
data HmmNode = HmmNode { transitions  :: TProbs
                       , matEmissions :: EProbs
                       , insEmissions :: EProbs
                       }
\end{smallverbatim}

>>>>

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
When performance is paramount,
software is usually written in C~or~C++. 
When convenience, readability, and
productivity are more important,
software is usually written in a dynamically typed
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on experience using a third kind of language,
Haskell: 
\begin{itemize}
\item
We had to reimplement an
algorithm already implemented in~C++, 
but the implementation required little effort,
and the connection between code and
algorithm is unusually clear (\secref{viterbi}).
Our new Haskell code performs slower than the~C++
implementation of the same algorithm,
but still meets our performance goals,
and it was easy to parallelize
(\secref{perf}).
However, this algorithm is only one part of a larger
algorithm, which on the whole performs significantly
better than the~C++ software that precedes it.
\item
Higher-order functions made it unusually easy to
create and experiment with new stochastic-search algorithms
(\secref{hofs}).
\item
Haskell slowed us down in only one way: it's hard to understand and
improve the
performance of a Haskell program (\secref{awkward-profiling}).
%%  \item
%%  The only significant impediment that Haskell presented to our
%%  research was the difficulty of understanding and improving the
%%  performance of Haskell programs (\secref{awkward-profiling}).
\item
Although authors Daniels and Gallant are computational
biologists with little background in functional programming,
we~used Haskell successfully.
And Haskell has made it easy to explore new research ideas;
by~contrast,
our group's C++ code has not been a good vehicle for exploring new ideas
(\secref{comparo}).
\item
The Haskell community offers libraries and tools that
promise powerful abstractions.
Some worked as advertised, saved us lots of effort, and were a pleasure
to~use.
Others, not so much.
And~we could not tell ahead of time which experience we were likely to have
 (\secref{penumbra}).
\end{itemize}

@@

In~idiomatic Haskell, 
we might represent an individual amino acid~$x$
using a value of algebraic data type:
\begin{smallverbatim}
data Amino = Ala | Cys | Asp | Glu | Gly | ...
\end{smallverbatim}
But our primary goal was to get something working quickly,
and our hidden Markov models are stored using a
legacy file format that represents each amino acid as a small integer.
Our representation is therefore
\begin{smallverbatim}
newtype AA = AA Int -- amino acid using HMMER coding
\end{smallverbatim}
Either way, a query sequence is (and should be) an immutable vector of
amino acids.
%type QuerySeq = Vector AA
%\end{smallverbatim}

@@

The representation of a node includes the probabilities of transitions
into the states of that node, plus the tables
of emission probabilities for the match and insertion states.
\remark{Why strictness annotations?}
\smallverbatiminput{hmmnode}
